P.berghei pir analysis

### 2021-09-13 chapter3_results_bergheiPir/Manuscript_Github_Pberghei_pir.Rmd
Remove unneeded package loads, get packages to work with ggplot2 3.3.3
Add error bars to smaller inset of stacked bar chart

@@

```{r ordering_vectors}
order_vector <- c('Sporo', 'Liv-pre4h', 'Liv-pre47h', 'Liv-post48h',
                 'Liv-2h','Liv-4h','Liv-12h', 'Liv-18h',
                 'Liv-24h' ,'Liv-36h','48h','54h','60h',
                 'Mero','Asex','Schiz','Gam','-Pre3h','BldMl-24h','Ook')

@@

```{r SorL_stacked_eachGene_fig2a}
  tpm_data <- combined_data %>% 
    rownames_to_column(var = 'Geneid') %>% 
    filter(Geneid %in% pir_id) %>% 
    melt(id.vars = 'Geneid', 
         variable.name = 'stage', 
         value.name = 'tpm') %>% 
  mutate(clade = pir_clade$clade[match(Geneid, 
                                       pir_clade$Geneid)]
         ) %>% 
  group_by(clade) %>% 
  mutate(clade = paste0(clade, ' (n = ', n_distinct(Geneid) ,')')) %>% 
    mutate(stack_group = Geneid) %>% 
    mutate(tpm = ifelse(tpm > 1, tpm, 0)) %>% #Filtering step for low value pir genes.
    group_by(stage, 
             stack_group,
             clade) %>% 
    dplyr::summarise(total_tpm = sum(tpm)) %>%
  arrange(
          desc(clade),
          total_tpm
)

tpm_data <- mutate(tpm_data,
                   order = row_number())

tpm_data_SorL_stackgenes <- mutate(tpm_data,
                        SorL = str_extract(clade, 
                                           pattern = '(L|S(?=[[:digit:]]))|ancestral|pan-rodent'))
tpm_data_SorL_stackgenes <- group_by(tpm_data_SorL_stackgenes,
                                     stage, 
                                     SorL,
                                     stack_group) %>% 
  dplyr::summarise(total_tpm = sum(total_tpm)) %>% 
  #Include the number of genes in each fill category for the labels
  mutate(SorL = paste0(SorL, 
                       ' (n = ', 
                       n_distinct(stack_group),
                       ')')) %>% 
#Order by SorL and then by total_tpm, then convert stack_group to a factor so ggplot orders correctly
  arrange(SorL, 
          total_tpm) %>% 
  mutate(stage_num = as.integer(stage))
tpm_data_SorL_stackgenes$stage %>% as.integer %>% unique
tpm_data_SorL_stackgenes$stage %>% unique

tpm_data_SorL_stackgenes$stack_group_order <- seq_len(length.out = nrow(tpm_data_SorL_stackgenes))
```

```{r stacked_bar_chart_figure2b}
#Calculate the range for the y value to make it look nicer
stacked_max <- tpm_data %>% 
  ungroup %>% 
  group_by(stage) %>% 
  dplyr::summarise(max_tpm = sum(total_tpm)) %>% 
  .$max_tpm %>% 
  max

@@

write_csv(x = tpm_data_SorL_stackgenes,
          path = 'tl20-05_SorL-subfamily_pir.csv')

stage_vec <- unique(
  levels(tpm_data$stage)
)

@@

y <- ggplot(data = tpm_data_SorL_stackgenes, 
            aes(
              x = stage_num,
              y = total_tpm, 
              fill = SorL,
              # group = stage_num
            )) +
  geom_rect(
    aes(
      xmin = stage_num - .7,
      xmax = stage_num + .7,
      ymin = -Inf,
      ymax = Inf,
      fill = stage_col
    ),
    alpha = .05
  ) +
  geom_bar(position = position_stack(),
           stat = 'identity',
           colour = 'black',
           size = 0.1) + 
  scale_fill_manual(
    breaks = unique(tpm_data_SorL_stackgenes$SorL),
    values = c(
      SorL_col_num,
      fills
    )
  ) +
  scale_color_manual('black') + 
  # labs(fill = stack_label) +
  scale_y_continuous(breaks = max_tpm) +
  theme_classic() +
  # scale_size(range = c(0,3)) +
  ylab(
    bquote('total '~italic(.('pir'))~' TPM')
  ) +
  scale_x_continuous(
    breaks = tpm_data_SorL_stackgenes$stage_num %>% unique(),
    labels = tpm_data_SorL_stackgenes$stage %>% unique(),
    expand = c(0, 0)
  ) +
    theme(
      axis.text.x = element_text(angle = 90, 
                                 vjust = .5,
                                 size = 10),
      axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.key.size = unit(3,'mm'),
    legend.key.width = unit(2,'mm'),
    ) +
    xlab('stage') + 
    facet_grid(.~stage, 
               scales = "free_x",
               space = "free_x") +
    scale_x_discrete(expand=c(0,0)) +
    theme(
      axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x = element_blank(),
      # axis.line.x = element_blank(),
      axis.text.y = element_text(size = 10),
      axis.title.y = element_text(size = 10),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10),
      legend.key.size = unit(3,'mm'),
      legend.key.width = unit(2,'mm'),
      strip.text.x = element_text(size=8, 
                                  face = 'bold', 
                                  angle = 90),
      strip.background = element_rect(
        color="black", 
        size=0.1, 
        linetype="solid"
      ),
      panel.spacing.x = unit(0, "null")
    )

g <- ggplot_gtable(ggplot_build(y))
strip_t <- which(grepl('strip-t', g$layout$name))
stage_vec <- unique(
  levels(tpm_data$stage)
)
fills <- case_when(str_detect(stage_vec, pattern = 'Asex') ~ 'coral1' ,
                   str_detect(stage_vec, pattern = 'Liv') ~ 'plum2',
                   str_detect(stage_vec, pattern = 'Gam') ~ 'lightskyblue1',
                   str_detect(stage_vec, pattern  = 'Bld|Sporo|Ook') ~ 'lightgreen')
k <- 1
for (i in strip_t) {
  j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
  g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
  k <- k+1
}

grid.newpage()
grid.draw(g)

@@

low_express <- tpm_data_SorL_stackgenes %>% 
  group_by(stage) %>% 
  summarise(sum_tpm = sum(total_tpm)) %>% 
  filter(sum_tpm < 350)

@@

y <- ggplot(data = tpm_data_SorL_stackgenes_low, 
            aes(
              x = stage,
              y = total_tpm, 
              fill = SorL,
              # group = stage_num
            )) +
  geom_bar(position = position_stack(),
           stat = 'identity',
           colour = 'black',
           size = 0.1) + 
  scale_fill_manual(
    breaks = unique(tpm_data_SorL_stackgenes$SorL),
    values = c(
      SorL_col_num,
      fills
    )
  ) +
  scale_color_manual('black') + 
  labs(fill = stack_label) +
  scale_y_continuous(breaks = max_tpm) +
  theme_classic() +
  scale_size(range = c(0,3)) +
  ylab(
    bquote('total '~italic(.('pir'))~' TPM')
  ) + 
  facet_grid(.~stage, 
             scales = "free_x",
             space = "free_x") +
  scale_x_discrete(expand=expand_scale(add=1)) +
  theme(
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x = element_blank(),
    # axis.line.x = element_blank(),
    axis.text.y = element_text(size = 15),
    axis.title.y = element_text(size = 8),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 5),
    legend.key.size = unit(3,'mm'),
    legend.key.width = unit(2,'mm'),
    strip.text.x = element_text(size=6, 
                                face = 'bold', 
                                angle = 90),
    strip.background = element_rect(
      color="black", 
      size=0.1, 
      linetype="solid"
    ),
    panel.spacing.x = unit(0, "null")
  )

>>>>

global_save_var <- FALSE

@@

```{r ordering_vectors}
order_vector <- c('Sporo', 'Liv-pre4h', 'Liv-pre47h', 'Liv-post48h',
                 'Liv-2h','Liv-4h','Liv-12h', 'Liv-18h',
                 'Liv-24h' ,'Liv-36h','48h','54h','60h',
                 'Mero','Asex','Ring','Troph','Schiz',
                 'Gam','Female','Male',
                 '-Pre3h','BldMl-24h','Ook')

@@

```{r SorL_stacked_eachGene_fig2a}
melted_tpm <- filter(berghei_tpm, 
                     Geneid %in% pir_id) %>% 
  mutate(clade = pir_clade$clade[match(Geneid, 
                                       pir_clade$Geneid)]
  ) %>% 
  mutate(SorL = str_extract(clade, 
                            pattern = '((L|S)(?=[[:digit:]]))|ancestral')) %>% 
  melt(id.vars = c('Geneid','clade', 'SorL'), 
       variable.name = 'sample', 
       value.name = 'tpm') %>% 
  group_by(clade) %>% 
  mutate(clade = paste0(clade, ' (n = ', n_distinct(Geneid) ,')')) %>% 
  group_by(SorL) %>% 
  mutate(
    SorL = paste0(SorL, 
                  ' (n = ', 
                  n_distinct(Geneid),
                  ')'),
    stage = str_extract(sample, 
                        pattern = '(?<=_).+(?=_)'))

```{r stacked_barchart_fun}
stacked_barchart <- function(melted_tpm,
                             threshold = 1,
                             stack_group_name = 'Geneid',
                             sd_bars = FALSE){
  #Have to use sym and !! in order to get mutate to work 
  # (converting the stack_group_name to a symbol and then unquoting in the mutate function)
  stack_group_name <- rlang::sym(paste(stack_group_name))

  tpm_data <- mutate(melted_tpm, 
                     stack_group = (!!stack_group_name)) %>% 
    mutate(tpm = ifelse(tpm > threshold, tpm, 0)) %>% #Filtering step for low value pir genes.
    group_by(stage, SorL, stack_group, sample) %>% 
    summarise(sum_tpm = sum(tpm)) %>% 
    group_by(SorL, stack_group, stage) %>%
    summarise(mean_sum_tpm = mean(sum_tpm),
              sd_sum_tpm = sd(sum_tpm)) %>% 
    arrange(desc(SorL),
            mean_sum_tpm) %>% 
    ungroup() %>% 
    mutate(order = row_number())

  tpm_data_SorL_stackgenes <- 
    #Order by SorL and then by mean_sum_tpm, then convert stack_group to a factor so ggplot orders correctly
    arrange(tpm_data,
            stage,
            desc(SorL), 
            mean_sum_tpm) %>% 
    mutate(stage = factor(stage,
                          levels = ordering(unique(tpm_data$stage),
                                            order_vector = r_order_vector)),
           stage_num = as.integer(
             factor(stage,
                    levels = ordering(unique(tpm_data$stage),
                                      order_vector = r_order_vector))
           )) %>% 
    group_by(stage) %>% 
    mutate(stacked_mean = cumsum(mean_sum_tpm),
           sd_upper = cumsum(mean_sum_tpm) + cumsum(sd_sum_tpm),
           sd_lower = ifelse(cumsum(mean_sum_tpm) - cumsum(sd_sum_tpm) < 0,
                             0, 
                             cumsum(mean_sum_tpm) - cumsum(sd_sum_tpm)))

  tpm_data_SorL_stackgenes$stack_group_order <- seq_len(length.out = nrow(tpm_data_SorL_stackgenes))

  stacked_max <- tpm_data %>% 
    ungroup %>% 
    group_by(stage) %>% 
    dplyr::summarise(max_tpm = sum(mean_sum_tpm)) %>% 
    .$max_tpm %>% 
    max

@@

@@

  # Note that this code does not work with ggplot 3.3.5, but it does work with 3.3.3. 
  #  Some issue with the colours for the geom_rect background.

  y <- ggplot(data = tpm_data_SorL_stackgenes, 
              aes(
                x = stage_num,
                y = mean_sum_tpm, 
                fill = SorL
              )) +
    geom_rect(
      aes(
        xmin = stage_num - .7,
        xmax = stage_num + .7,
        ymin = -Inf,
        ymax = Inf,
        fill = stage_col
      ),
      alpha = .5
    ) +
    geom_bar(position = position_stack(),
             stat = 'identity',
             colour = 'black',
             size = 0.1) +
    {if(sd_bars)list(
      geom_errorbar(aes(ymax = sd_upper,
                        ymin = sd_lower),
                    position = position_dodge(width = .85),
                    size = .25),
        geom_point(aes(y = stacked_mean), 
                   position = position_dodge(width = .85),
                   show.legend = FALSE,
                   size = .25)
    )}+ 
    scale_fill_manual(
      breaks = unique(tpm_data_SorL_stackgenes$SorL),
      values = c(
        SorL_col_num,
        fills
      )
    ) +
    scale_color_manual('black') + 
    scale_y_continuous(breaks = max_tpm) +
    theme_classic() +
    ylab(
      bquote('total '~italic(.('pir'))~' TPM')
    ) +
    scale_x_continuous(
      breaks = tpm_data_SorL_stackgenes$stage_num %>% unique(),
      labels = tpm_data_SorL_stackgenes$stage %>% unique(),
      expand = c(0, 0)
    ) +
    theme(
      axis.text.x = element_text(angle = 90, 
                                 vjust = .5,
                                 size = 10),
      axis.text.y = element_text(size = 10),
      axis.title.y = element_text(size = 10),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10),
      legend.key.size = unit(3,'mm'),
      legend.key.width = unit(2,'mm'),
    ) +
    xlab('stage') + 
    facet_grid(.~stage, 
               scales = "free_x",
               space = "free_x") +
    scale_x_discrete(expand=c(0,0)) +
    theme(
      axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(size = 10),
      axis.title.y = element_text(size = 10),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10),
      legend.key.size = unit(3,'mm'),
      legend.key.width = unit(2,'mm'),
      strip.text.x = element_text(size=8, 
                                  face = 'bold', 
                                  angle = 90),
      strip.background = element_rect(
        color="black", 
        size=0.1, 
        linetype="solid"
      ),
      panel.spacing.x = unit(0, "null")
    )

  g <- ggplot_gtable(ggplot_build(y))
  strip_t <- which(grepl('strip-t', g$layout$name))
  stage_vec_strip <- sort(unique(tpm_data_SorL_stackgenes$stage))
  fills <- case_when(str_detect(stage_vec_strip, pattern = 'Asex') ~ 'coral1' ,
                     str_detect(stage_vec_strip, pattern = 'Liv') ~ 'plum2',
                     str_detect(stage_vec_strip, pattern = 'Gam') ~ 'lightskyblue1',
                     str_detect(stage_vec_strip, pattern  = 'Bld|Sporo|Ook') ~ 'lightgreen')
  k <- 1
  for (i in strip_t) {
    j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
    g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
    k <- k+1
  }

  grid.newpage()
  grid.draw(g)
  return(g)
}
```

```{r stacked_barchart}
sd_bars <- TRUE
g <- stacked_barchart(melted_tpm,
                      stack_group_name = 'SorL',
                      sd_bars = sd_bars)

if(global_save_var){
ggsave('plots/thesis_chapter3_figure_stackedBarChartTPM_SorL_standevError.png',
       width = 7,
       height = 5,
       units = "in",
       g)
}

@@

low_express <- tpm_data_SorL_stackgenes %>% 
  group_by(stage) %>% 
  summarise(sum_tpm = sum(mean_sum_tpm)) %>% 
  filter(sum_tpm < 350)

@@

y <- ggplot(data = tpm_data_SorL_stackgenes_low, 
              aes(
                x = stage_num,
                y = mean_sum_tpm, 
                fill = SorL
              )) +
    geom_bar(position = position_stack(),
             stat = 'identity',
             colour = 'black',
             size = 0.1) +
    {if(sd_bars)list(
      geom_errorbar(aes(ymax = sd_upper,
                        ymin = sd_lower),
                    position = position_dodge(width = .85),
                    size = .25),
        geom_point(aes(y = stacked_mean), 
                   position = position_dodge(width = .85),
                   show.legend = FALSE,
                   size = .25)
    )}+ 
    scale_fill_manual(
      breaks = unique(tpm_data_SorL_stackgenes$SorL),
      values = c(
        SorL_col_num,
        fills
      )
    ) +
    scale_color_manual('black') + 
    scale_y_continuous(breaks = max_tpm) +
    theme_classic() +
    ylab(
      bquote('total '~italic(.('pir'))~' TPM')
    ) +
    scale_x_continuous(
      breaks = tpm_data_SorL_stackgenes$stage_num %>% unique(),
      labels = tpm_data_SorL_stackgenes$stage %>% unique(),
      expand = c(0, 0)
    ) +
    theme(
      axis.text.x = element_text(angle = 90, 
                                 vjust = .5,
                                 size = 10),
      axis.text.y = element_text(size = 10),
      axis.title.y = element_text(size = 10),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10),
      legend.key.size = unit(3,'mm'),
      legend.key.width = unit(2,'mm'),
    ) +
    xlab('stage') + 
    facet_grid(.~stage, 
               scales = "free_x",
               space = "free_x") +
      scale_x_discrete(expand=expand_scale(add=1)) +
    theme(
      axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(size = 10),
      axis.title.y = element_text(size = 10),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10),
      legend.key.size = unit(3,'mm'),
      legend.key.width = unit(2,'mm'),
      strip.text.x = element_text(size=8, 
                                  face = 'bold', 
                                  angle = 90),
      strip.background = element_rect(
        color="black", 
        size=0.1, 
        linetype="solid"
      ),
      panel.spacing.x = unit(0, "null")
    )
